---
title: "Council accounts"
output: html_notebook
---

# Council accounts: comparing estimates vs spend

This notebook details the process of downloading, cleaning, and analysing council accounts.

The data comes from the Ministry of Housing, Communities & Local Government's [Local authority revenue expenditure and financing collection](https://www.gov.uk/government/collections/local-authority-revenue-expenditure-and-financing)

## Spend vs budget

Start with the [revenue outturn summary](https://www.gov.uk/government/statistics/local-authority-revenue-expenditure-and-financing-england-2017-to-2018-individual-local-authority-data-outturn). 

The [specific file is here](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/735627/RS_2017-18_data_by_LA.xlsx). Sheet RS LA Data 2017-18, columns H and I: Children social care and adult social care respectively.

Then compare with the [budget](https://www.gov.uk/government/statistics/local-authority-revenue-expenditure-and-financing-england-2017-to-2018-budget-individual-local-authority-data). Specific [spreadsheet here](https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/623097/RA_2017-18_data_by_LA.xlsx). Sheet RA LA Data 2017-18 - Columns AK (Children’s social care) and BD (Adults).

Then the difference between them.

And then repeat for previous years.

[2016-17 outturn](https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/659764/RS_2016-17_data_by_LA.xlsx): Children’s in column H, Adults in column I. 

[2016-17 budget](https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/532962/RA_2016-17_data_by_LA.xlsx): Children’s in AJ, Adults in BC

[2015-16 outturn](https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/659783/RS_2015-16_data_by_LA_-_Revision.xlsx): Children’s in column H, Adults in column I


[2015-16 budget](https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/444910/RA_2015-16_data_by_LA_-_Nat_Stats_Release_-_15-Jul-2015.xls): Children’s in Column AJ, Adults in Column BC

Let's store those URLs:

```{r}
outturn1718url <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/735627/RS_2017-18_data_by_LA.xlsx"
budget1718url <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/623097/RA_2017-18_data_by_LA.xlsx"
outturn1617url <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/659764/RS_2016-17_data_by_LA.xlsx"
budget1617url <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/532962/RA_2016-17_data_by_LA.xlsx"
outturn1516url <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/659783/RS_2015-16_data_by_LA_-_Revision.xlsx"
budget1516url <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/444910/RA_2015-16_data_by_LA_-_Nat_Stats_Release_-_15-Jul-2015.xls"
cellrange.outturn <- "H:I"
cellrange.budget <- "AJ:BC"
```


## Breaking down the problem

Here are the tasks we need to complete:

1. Download the spreadsheet directly from a URL
2. Perform basic cleaning (remove redundant rows, combine/clarify headings across multiple cells)
3. Create a subset containing the data we want to focus on
4. Perform analysis

## Download the data - 1617

```{r}
#Activate the library for handling Excel files, and for calling URLs
library(readxl)
library(httr)
#For 2016-17:
#Fetch that file and save it locally with a new name
GET(outturn1617url, write_disk("revenueaccounts1617.xlsx", overwrite = T))
GET(budget1617url, write_disk("budgetaccounts1617.xlsx", overwrite = T))
```

## Clean the spreadsheet while importing

Now to import that, and clean it along the way. There are two problems here: first, the column headers don't start until row 4; and second, those headers are spread across 3 rows.

The most detailed headings are in row 7, with related codes for each heading in row 6, and broader categories in row 5. So we skip to row 7 for the main dataset...

```{r}
#Activate the dplyr library
library(dplyr)
#Read that file into a new data frame
#We want the 3rd sheet, and to skip the first 6 lines so row 7 is used for headers
revenueaccounts1617 <- read_excel("revenueaccounts1617.xlsx", sheet = 3, skip = 6, na = "c('-','')") #could add range = cell_cols("A:I") but skip won't work
budgetaccounts1617 <- read_excel("budgetaccounts1617.xlsx", sheet = 3, skip = 6, na = "c('-','')") #omitted: range = cell_cols(cellrange.budget)
```

...and store the other headings in another data frame just in case.

```{r}
#Rows 5 and 6 also contain useful info so we grab those just in case
#We specify the range because otherwise it skips the blank cells in the first 5 columns
#An alternative would be to change or remove the n_max so there are some full cells in those first 5 cols
revenueaccounts1617.otherheads <- read_excel("revenueaccounts1617.xlsx", sheet = 3, skip = 4, n_max = 2) #we set n_max to 2 because otherwise it skips the first 5 empty cells
#Most of these are empty so we need to copy from the previous cell.
#Start from index 2 because the i-1 below will generate an error otherwise
for(i in seq(2,length(revenueaccounts1617.otherheads))){
  #grab first two characters
  first2 <- substr(colnames(revenueaccounts1617.otherheads)[i],0,2)
  #If this is an unnamed column
  if (first2 == "X_"){
    #Replace with the name of the previous column
    colnames(revenueaccounts1617.otherheads)[i] <- colnames(revenueaccounts1617.otherheads)[i-1]
  }
}
```


## Create a subset

There are 247 variables in one spreadsheet, and 99 in the other, so let's narrow down to the ones we want to focus on.

We can use regex to find the columns where 'Social' is mentioned:

```{r}
#The grepl function is used to generate a series of TRUE/FALSE matches for each column
#That is then nested within square brackets after the colnames function to show those that are TRUE
#The (?i) modifier at the start of the regex specifies that we want to be case insensitive, which means we can catch the capital/lower case S at the start and where it is all caps.
colnames(revenueaccounts1617)[grepl("(?i).*social.*",colnames(revenueaccounts1617))]
colnames(budgetaccounts1617)[grepl("(?i).*social.*",colnames(budgetaccounts1617))]
```


Now that we think we have the right columns, let's generate that subset:

```{r}
#We remove 'colnames' now so we are accessing the data frame as a whole
revenueaccounts1617.social <- revenueaccounts1617[grepl("(?i).*social.*",colnames(revenueaccounts1617))]
#Show the relative otherheads too
revenueaccounts1617.otherheads[grepl("(?i).*social.*",colnames(revenueaccounts1617))]
revenueaccounts1617.social.otherheads <- revenueaccounts1617.otherheads[grepl("(?i).*social.*",colnames(revenueaccounts1617))]

#And for budget
budgetaccounts1617.social <- budgetaccounts1617[grepl("(?i).*social.*",colnames(budgetaccounts1617))]
```

And combine with the basic data on name, code etc.

```{r}
#Create a subset of just the basic names and codes in the first few columns - we'll probably need this again so useful to save as a separate data frame
councillist <- revenueaccounts1617[c(1:3,5)]
councillist.ba <- budgetaccounts1617[c(1:2,4)]
#Combine the two using cbind
revenueaccounts1617.social <- cbind(councillist,revenueaccounts1617.social)
budgetaccounts1617.social <- cbind(councillist.ba,budgetaccounts1617.social)
```

We've now gone from 247 variables to 6, which is going to be much easier to deal with.

### Subsetting by type of organisation - revenue accounts

We also need to remove all the organisations we don't want. The type of organisation is shown in the 'Class' column:

```{r}
table(revenueaccounts1617.social$Class)
```

We don't want shire districts (SD), or other authorities such as police or fire (O). 

```{r}
revenueaccounts1617.social <- subset(revenueaccounts1617.social, revenueaccounts1617.social$Class != "O" & revenueaccounts1617.social$Class != "SD")
table(revenueaccounts1617.social$Class)
```

This gives us a data frame with 156 rows - 4 more than we might expect. This is because aggregate figures for each type (4 types) are also in the data. 

These make up the last 4 rows and have no code:

```{r}
#Subset so we only have rows where the code is not NA
revenueaccounts1617.social <- subset(revenueaccounts1617.social,revenueaccounts1617.social$`E-code` != "NA")
#Check a table to see each number has gone down by 1
table(revenueaccounts1617.social$Class)
```



## Cleaning data types

To check if we need more cleaning we can generate a summary:

```{r}
summary(revenueaccounts1617.social)
```

This indicates that many of the columns have been imported as characters, rather than numeric, most likely because of the presence of dashes.

We need to fix this. Let's use the `tidyverse` library and use its `guess_parser` function to see what it thinks of one of the columns:

```{r}
library(tidyverse)
#Guess the first number columns
guess_parser(revenueaccounts1617.social[,5])
```

Now a `summary` to see much the same:

```{r}
summary(revenueaccounts1617.social[,6])
```

We'll need to parse it instead as a number during any analysis.

```{r}
#Replace column with numeric version
revenueaccounts1617.social[,6] <- parse_number(revenueaccounts1617.social[,6])
#Sum - this returns NA because we need to subset out the NA
sum(revenueaccounts1617.social[,6])
#Table to show frequency of numbers
table(revenueaccounts1617.social[,6])
summary(revenueaccounts1617.social[,6])
```

Repeat for the other column:

```{r}
#Replace column with numeric version
revenueaccounts1617.social[,5] <- parse_number(revenueaccounts1617.social[,5])
#Sum - this returns NA because we need to subset out the NA
sum(revenueaccounts1617.social[,5])
#Table to show frequency of numbers
table(revenueaccounts1617.social[,5])
summary(revenueaccounts1617.social[,5])
```

The presence of an NA in the data (Isles of Scilly) prevents us from calculating a sum- We need to subset out the NA to do this.

```{r}
forsum <- subset(revenueaccounts1617.social, revenueaccounts1617.social$`Children Social Care` != "NA")
sum(forsum$`Children Social Care`)
#Or we can combine the two, removing the need to create a new object
sum(subset(revenueaccounts1617.social, revenueaccounts1617.social$`Children Social Care` != "NA")$`Children Social Care`)
```

```{r}
#Export a pure R script version of this notebook
knitr::purl("3estimate_vs_spend.Rmd", "3estimate_vs_spend.R", documentation = 2)
```


