---
title: "Council accounts: resilience test"
output: html_notebook
---

# Testing resilience by analysing council accounts

This notebook details the process of analysing council accounts in order to arrive at a measure of 'resilience'.

The data comes from the Ministry of Housing, Communities & Local Government's [Local authority revenue expenditure and financing collection](https://www.gov.uk/government/collections/local-authority-revenue-expenditure-and-financing#2018-to-2019). 

We can find the latest by going to [Local authority revenue expenditure and financing collection](https://www.gov.uk/government/collections/local-authority-revenue-expenditure-and-financing#2018-to-2019) and looking for the latest "Local authority revenue expenditure and financing England: ... individual local authority data - outturn" then within that "Revenue outturn summary (RS) ...".

Begin by changing the line below so that the URL specifies the latest **Revenue account (RA) budget** URL:

```{r}
#Update this when new data is available
rourl <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/740771/RS_2017-18_data_by_LA.xlsx"
```


## Methodology

The methodology is adapted from that detailed in the [CIPFA consultation on an index of resilience](https://www.cipfa.org/policy-and-guidance/consultations/consultation-on-cipfa-index-of-resilience-for-english-councils). This identifies the following measures:

* The level of total reserves excluding schools and public health as a proportion of net revenue expenditure (from RO outturns) - weighted 0.25
* The percentage change in reserves, excluding schools and public health, over the past three years (from RO outturns) - weighted 0.25
* The ratio of government grants to net revenue expenditure (from RO outturns)  - weighted 0.1
* Proportion of net revenue expenditure accounted for by children’s social care, adult social care and debt interest payments (from RO outturns) - weighted 0.15
* Ofsted overall rating for children’s social care ([Ofsted report, released in July annually](https://www.gov.uk/government/statistics/childrens-social-care-data-in-england-2018)) - weighted 0.15
* Auditor’s VFM judgement (Public Sector Audit Appointment’s summary of the results of audit: an adverse or “except for” judgement may be indicative of poor financial management within a council) - weighted 0.1

These notebooks focus on the first four measures from RO outturns.

## The level of total reserves excluding schools and public health as a proportion of net revenue expenditure

Let's grab this from the latest data - using `raurl` defined at the start of this notebook:


```{r}
#Activate the library for calling URLs
library(httr)
#Fetch that file and save it locally with a new name
GET(rourl, write_disk("revenueoutturn.xlsx", overwrite = T))
```

Now to import that, and clean it along the way. There are two problems here: first, the column headers don't start until row 4; and second, those headers are spread across 3 rows.

The most detailed headings are in row 7, with related codes for each heading in row 6, and broader categories in row 5. So we skip to row 7 for the main dataset, and store the other headings in another data frame just in case.

```{r}
#Import the library we need to read Excel
library(readxl)
#Read that file into a new data frame
#We want the 3rd sheet, and to skip the first 6 lines so row 7 is used for headers
rodata <- read_excel("revenueoutturn.xlsx", sheet = 3, skip = 6, na = "c('-','')")
#Rows 5 and 6 also contain useful info so we grab those just in case
#We specify the range because otherwise it skips the blank cells in the first 5 columns
#An alternative would be to change or remove the n_max so there are some full cells in those first 5 cols
ro.otherheads <- read_excel("revenueoutturn.xlsx", sheet = 3, skip = 4, n_max = 1, range = "A5:IR6")
#Most of these are empty so we need to copy from the previous cell.
#Start from index 2 because the i-1 below will generate an error otherwise
for(i in seq(2,length(ro.otherheads))){
  #grab first two characters
  first2 <- substr(colnames(ro.otherheads)[i],0,2)
  #If this is an unnamed column
  if (first2 == "X_"){
    #Replace with the name of the previous column
    colnames(ro.otherheads)[i] <- colnames(ro.otherheads)[i-1]
  }
}
```


### Find the data points we need

There are 100 variables in this spreadsheet, so let's narrow down to the ones we want to focus on.

* "The level of total reserves" 
* "excluding schools and public health" 
* "as a proportion of"
* "net revenue expenditure"

We can use regex to find the columns where 'reserves' are mentioned:

```{r}
colnames(rodata)[grepl(".*[Rr]eserves.*",colnames(rodata))]
```

And narrow down further by simply adding a space - this excludes the ones that *end* with "reserves" and no space after:

```{r}
colnames(rodata)[grepl(".*[Rr]eserves .*",colnames(rodata))]
```

Now that we think we have the right columns, let's generate that subset:

```{r}
#We remove 'colnames' now so we are accessing the data frame as a whole
rodata.res <- rodata[grepl(".*[Rr]eserves .*",colnames(rodata))]
#Create a subset of just the basic names and codes in the first few columns - we'll probably need this again so useful to save as a separate data frame
rodata.councils <- rodata[c(1:3,5)]
#Combine the two using cbind
rodata.res <- cbind(rodata.councils,rodata.res)
```

We've now gone from 100 variables to 14, which is going to be much easier to deal with.

### Subsetting by type of organisation

We also need to remove all the organisations we don't want. The type of organisation is shown in the 'Class' column:

```{r}
table(rodata.res$Class)
```

We don't want shire districts (SD), or other authorities such as police or fire (O). 

```{r}
rodata.res <- subset(rodata.res, rodata.res$Class != "O" & rodata.res$Class != "SD")
table(rodata.res$Class)
```

This gives us a data frame with 156 rows - 4 more than we might expect. This is because aggregate figures for each type (4 types) are also in the data. 

These make up the last 4 rows and have no code:

```{r}
#Subset so we only have rows where the code is not NA
rodata.res <- subset(rodata.res,rodata.res$`E-code` != "NA")
#Check a table to see each number has gone down by 1
table(rodata.res$Class)
```



## Cleaning data types

To check if we need more cleaning we can generate a summary:

```{r}
summary(rodata.res)
```

This indicates that many of the columns have been imported as characters, rather than numeric, most likely because of the presence of dashes.

We need to fix this. Let's use the `tidyverse` library and use its `guess_parser` function to see what it thinks of one of the columns:

```{r}
library(tidyverse)
#Guess the first number columns
guess_parser(rodata.res[,5])
```

Now a `table` to see what values are in that column:

```{r}
summary(rodata.res[,5])
table(rodata.res[,5])
```

We'll need to parse it instead as a number during any analysis.

```{r}
#Parse as number
parse_number(rodata.res[,5])
#Replace column with numeric version
rodata.res[,5] <- parse_number(rodata.res[,5])
#Sum - this returns NA
sum(rodata.res[,5])
#Table to show frequency of numbers
table(rodata.res[,5])
```

Here's a quicker way to do this for all the columns from 5 onwards

```{r}
#Create a for loop to go through each number from 5 to the length of rodata.res
for(i in seq(5,length(rodata.res)))
  {
  #Replace the column at that position with the same column parsed as numbers
  rodata.res[,i] <- parse_number(rodata.res[,i])
  }
#Check the resulting dataframe
summary(rodata.res)
```

All of these have 2 NA entries, so let's look at those.

```{r}
is.na(rodata.res$`Estimated schools reserves level at 1 April`)
```

```{r}
#Export a pure R script version of this notebook
knitr::purl("4resiliencetest.Rmd", "4resiliencetest.R", documentation = 2)
```